Extracting The Power of Dependent Types

### 2021-07-13 paper.tex
Comment out triangle picture for now
Clarify itnroduction.

When writing a program, it is often desirable to guarantee that a
certain class of errors does not occur in the given program.  For
example, the program can never dereference a null pointer; or division
by zero never occurs; or indexing is always within bounds.  In most
programming languages, the class of guarantees that can be checked
statically is very limited.  Nevertheless, moving to a different
language is not always an option.  For example, these languages may
come with great tooling, compilers and libraries.  In this paper, we
investigate how to provide strong static guarantees for programs in a
given language, while maintaining the ability to to use existing
compilers and tools.

@@

While it is possible to add dependent types to an existing language
from scratch, doing so is time-consuming and error-prone.  Instead,
one can also embed the language we want to use into an existing
dependently typed language, which allows us to re-use the typechecker
of the host language to typecheck the embedded language.

There are two common approaches to language embedding: deep and shallow.
A shallow embedding defines a language in terms of functions and operators
of the host language.  Programs in a shallowly-embedded language are programs
in the host language, so they are immediately runnable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs.  The downside of this approach is that we cannot
easily get the representation of the embedded program.

Meanwhile, a deep embedding defines a data type to represent the AST
of the embedded language.  This requires us to define a type system
and evaluator for the language separately.  With a dependently typed
host language the AST and the type system can be defined as a single
data structure, which is typically referred to as a tagless
interpreter~\cite{}.  The benefit of deep embedding is that we get
full access to the syntactic structure of the embedded program.  This
approach works beautifully for embedded languages with simple type
systems. However, for embedded languages with dependent types,
defining a tagless interpreter is technically possible~\cite{}, but
extremely challenging to define and use in practice.

We propose a novel approach that combines the best parts of shallow
and deep embeddings through the use of reflection in a
dependently-typed host language such as Coq, Agda, Idris, or Lean.
Concretely, we start by defining a shallow embedding of the target
language while making use of dependent types to statically enforce the
properties of interest, and write a program in this embedding. We then
make use of reflection to obtain a syntactic representation of the
program, and extract a program in the syntax of the target language
from it.  Finally, we compile the extracted program using existing
tools and libraries. Essentially, a shallow embedding in a host
language with reflections suddenly becomes deep.

In this paper we demonstrate that the proposed approach is actually possible
by example.  We embed a subset of PostScript~\cite{}, which is a stack-based
language used in the corresponding document format.  The simplicity of the language
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We use Agda as our host language, and we demonstrate the
embedding, how to specify a program, and most importantly, program extraction.

\paragraph{Limitations}

While we claim our approach is flexible and powerful, it is
certainly not a magic bullet. There are still a lot of things that
need to be implemented for each target language: the language
embedding itself, the encoding of the properties under consideration
as types in the host language, and the extractor that transforms
embedded programs back into the syntax of the original language.  For
enforcing more complex properties, it might also be necessary to write
proofs to convince the typechecker of the host language that they are
indeed satisfied.
%
The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.  In the
end, our goal is not to showcase a single example but to empower you
to apply this approach yourself to verify code in any language.

@@

\item We demonstate this approach by a concrete example: an embedding
  of (a subset of the) PostScript language into Agda.

@@

\todo[inline]{Move it somewhere}
This picture has been generated by the code that was extracted from our
Agda embedding.

\begin{center}
\epsfbox[0 0 64 64]{sierp.ps}
\end{center}

@@

>>>>

It is often desirable to guarantee that a
certain class of errors does not occur in the given program.  These
error classes may include: dereferencing the null pointer, division
by zero, or out of bound indexing.  In most
programming languages, the guarantees that can be checked
statically are very limited.  Nevertheless, moving to a different
language is not always an option.  For example, these languages may
come with great tooling, compilers and/or libraries.  In this paper, we
investigate how to provide strong static guarantees for programs in a
given language, while maintaining the ability to to use existing
compilers and tools.

@@

While it is possible to add dependent types to an existing language
from scratch, doing so is time-consuming and error-prone.  Instead,
one can also embed the language we want to use into existing
dependently typed languages, which allows us to re-use the typechecker
of the host language to typecheck the embedded language.

There are two common approaches to language embedding: deep and shallow.
A shallow embedding defines a language in terms of functions and operators
of the host language.  Programs in a shallowly-embedded language are programs
in the host language, so they are immediately runnable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs.  The downside of this approach is that we cannot
easily get the representation of embedded programs.

Meanwhile, a deep embedding defines a data type to represent the AST
of the embedded language.  This requires us to define a type system
and evaluator for the language separately.  With a dependently typed
host language the AST and the type system can be defined as a single
data structure, which is typically referred to as a tagless
interpreter~\cite{}.  The benefit of deep embedding is that we get
full access to the syntactic structure of the embedded program.  This
approach works beautifully for embedded languages with simple type
systems. However, for embedded languages with dependent types,
tagless interpreters are technically possible~\cite{}, but
are extremely challenging to define and use in practice.

The key insight of this paper lies in proposing a novel approach that
combines the best parts of shallow and deep embeddings through the use
of reflection in a
dependently-typed host language such as Coq, Agda, Idris, or Lean.

Our main objective of the paper is to demonstrate that the
proposed approach works in practice.  Therefore, we guide the reader
through the entire process of: i) shallow embedding of the target
language;  ii) writing programs in this embedding; iii) using
dependent types to statically enforce the properties of interest; 
iv) extract a program in the syntax of the target language
from it.  We run the extracted program using existing
tools and libraries. Essentially, demonstrating that a
shallow embedding in a host language with reflections
suddenly becomes deep.

In order to keep presentation as simple as possible, we have
chosen a very simple target language to embed --- we use a subset
of PostScript~\cite{}, which is a stack-based
language used in the corresponding document format.  While the embedding
may look impractical, the simplicity of the language
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, implying that the proposed approach scales.
We use Agda as our host language, and the entire paper is Agda script.
This means that the code presented in the snippets has been typechecked
and presented examples ran.

@@

\item We demonstrate this approach by a concrete example: an embedding
  of (a subset of the) PostScript language into Agda.

@@

\begin{comment}
\todo[inline]{Move it somewhere}
This picture has been generated by the code that was extracted from our
Agda embedding.

\begin{center}
\epsfbox[0 0 64 64]{sierp.ps}
\end{center}
\end{comment}

@@

\paragraph{Limitations}
\todo[inline]{This has moved from the intro, needs reflowing!}
While we claim our approach is flexible and powerful, it is
certainly not a magic bullet. There are still a lot of things that
need to be implemented for each target language: the language
embedding itself, the encoding of the properties under consideration
as types in the host language, and the extractor that transforms
embedded programs back into the syntax of the original language.  For
enforcing more complex properties, it might also be necessary to write
proofs to convince the typechecker of the host language that they are
indeed satisfied.
%
The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.  In the
end, our goal is not to showcase a single example but to empower you
to apply this approach yourself to verify code in any language.
